---
title: "Homework 5"
author: "Shivam Bajaj"
date: "2/13/2023"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

<style>div.soln { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}</style>

```{r setup, include=FALSE, echo=T}
knitr::opts_chunk$set(echo = F, message = F, warning = F, collapse = T)
library(tidyverse)
```


***
1. Using only `rnorm`, draw a sample of 1000 random numbers from $\chi^2_7$. Plot a histogram. **1 point**

```{r}
sample <- rchisq(n = 1000, df = 7)
hist(sample, breaks = 20, main = "Hist of 1000 random no. Chi-squared distribution with 7 DoF")

```

***
2. Plot the log-likelihood of the sample assuming a $\chi^2_{df}$ distribution under candidate `df` values from 1 to 14. **1 point**
```{r}

loglik_chi <- function(df, x){
  sum(dchisq(x,df=df, log=TRUE))
}

sample_loglik <- sapply(1:14,function(df) loglik_chi(df,sample))

plot(1:14, sample_loglik, type = "l", xlab = "df", ylab = "Log-likelihood",
     main = "Log-likelihood of sample assuming Chi-squared distribution")
```

***
3. Using only `rnorm` (and potentially your answer from question 1) draw a sample of 1000 random numbers from $t_7$. Plot a histogram. **1 point**
```{r}
chi_sq_sample <- rchisq(n = 1000, df = 7)
z <- rnorm(n = 1000)

#t-distribution with 7 degrees of freedom
x <- sqrt(7) * z / sqrt(chi_sq_sample)

hist(x, breaks = 20, main = "Hist. of 1000 random no. from t-distribution with 7 DoF")
```

***
4. Plot the log-likelihood of your sample assuming a $t_{df}$ distribution under candidate `df` values from 1 to 14. **1 point**
```{r}
loglik_t <- function(df, x) {
  sum(dt(x, df = df, log = TRUE))
}

sample_loglik_t <- sapply(1:14, function(df) loglik_t(df, x))

plot(1:14, sample_loglik_t, type = "l", xlab = "df", ylab = "Log-likelihood",
     main = "Log-likelihood of sample assuming t-dist.")
```

***
5. Using only `rnorm` (or potentially your answer from question 1), draw a sample of 1000 random numbers from $F_{4,4}$. Plot a histogram. **1 point**. 
```{r}
z1 <- rnorm(n = 1000)
z2 <- rnorm(n = 1000)

chi_sq1 <- rchisq(n = 1000, df = 4)
chi_sq2 <- rchisq(n = 1000, df = 4)

f_sample <- (z1 / sqrt(chi_sq1 / 4)) / (z2 / sqrt(chi_sq2 / 4))
f_sample <- f_sample / (7 / chi_sq1 + 7 / chi_sq2)

hist(f_sample, breaks = 20, main = "Hist. of 1000 random numbers from F-dist. with 4 and 4 DoF")

```

***
6. Plot the log-likelihood of your sample assuming a $F_{df_1,df_2}$ distribution with $df_1$ fixed at 4 under candidate `df_2` values from 1 to 14. **1 point**
```{r}
loglik_f <- function(df2, x) {
  sum(df2/2 * log(df2/2) + (df2/2 + 2) * log(x) -
        (df2/2 + df2 * x / 2))
}

f_sample <- rf(n = 1000, df1 = 4, df2 = 4)

sample_loglik_f <- sapply(1:14, function(df2) loglik_f(df2, f_sample))

plot(1:14, sample_loglik_f, type = "l", xlab = "df2", ylab = "Log-likelihood",
     main = "Log-lik. of sample assuming F-dist. with fixed df1 ")
```

***
7. Write a short function to convert a test statistic $x$ into the area under the $z$-density for values $≥|x|$. This area is the P value. The function should take $x$ as an argument. **(1 point)**
```{r}

p_value <- function(x) {
  p <- 1 - pnorm(abs(x))
  return(p)
}

```

***
8. Write a short function to convert a test statistic $x$ into the area under the $t_{df}$-density for values $≥|x|$. This area is the P value. The function should take $x$ and $df$ as arguments. **(1 point)**
```{r}
p_value_t <- function(x, df) {
  p <- 1 - pt(abs(x), df)
  return(p)
}
```

***
9. Simulate the results of a cancer trial.

Let's say that in a double-blind randomized clinical breast cancer trial, the primary outcome is *tumor diameter*.
At study end, the average tumor reduction in tumor size was 3 ± 1 cm (mean ± sd) in the treatment group and 0 ± 1 cm (mean ± sd) in the control group.
Simulate reduction values for N = 12 treatment and N = 12 control participants. Use `t.test` to whether the reductions differ between groups.
Compare the results when using the default options and when using `var.equal = T`. **(1 point)**
```{r}
treatment <- rnorm(n=12,mean=3,sd=1)

control <- rnorm(n=12,mean=0,sd=1)

t_test_default <- t.test(treatment,control)

t_test_equal <- t.test(treatment, control, var.equal= TRUE)

print(t_test_default)

print(t_test_equal)

```

***
10. Recreate the functionality of `t.test`. Use base R functions (e.g., `length`, `mean`, `var`, `sqrt`) and `pt` to recreate the t statistic and p value from problem 9 without `t.test` using `var.equal=TRUE`. **(1 point)**
```{r}
treatment <- rnorm(n = 12, mean = 3, sd = 1)
control <- rnorm(n = 12, mean = 0, sd = 1)


x1 <- mean(treatment)
x2 <- mean(control)
s1 <- sd(treatment)
s2 <- sd(control)

t_test <- function(x1,x2,s1,s2,n1,n2){
  t_stat <- (x1-x2) / sqrt((s1^2/n1)+(s2^2/n2))
  df <- (s1^2 / n1 + s2^2 / n2)^2 / ((s1^2 / n1)^2 / (n1 - 1) + (s2^2 / n2)^2 / (n2 -1))
  p_val <- 2 * pt(abs(t_stat), df, lower.tail = FALSE)
  return(list(t_stat = t_stat, df = df, p_val = p_val))
}

t_test_equal <- function(x1,x2,s1,s2,n1,n2){
  s_p <- sqrt(((n1 - 1) * s1^2 + (n2 - 1) * s2^2) / (n1 + n2 - 2))
  t_stat <- (x1 - x2) / (s_p * sqrt(1/n1 + 1/n2))
  df <- n1 + n2 - 2
  p_val <- 2 * pt(abs(t_stat), df, lower.tail = FALSE)
  return(list(t_stat = t_stat, df = df, p_val = p_val))
}

results <- t_test(x1, x2, s1, s2, length(treatment), length(control))
cat("unequal var\n")
cat("t-statistic:", results$t_stat, "\n")
cat("degrees of freedom:", results$df, "\n")
cat("p-value:", results$p_val, "\n")


results <- t_test_equal(x1, x2, s1, s2, length(treatment), length(control))
cat("equal var\n")
cat("t-statistic:", results$t_stat, "\n")
cat("degrees of freedom:", results$df, "\n")
cat("p-value:", results$p_val, "\n")


```