---
title: "Data Wrangling"
author: "J. Lucas McKay"
date: "9/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
```

## Filling omitted repetitions

In one common output format, repeated values are omitted, and are only recorded when they change.
Let's download some deidentified gait data and have a look.
These are actual data; however, the dates have been rotated by a random amount for each patient to
protect privacy (while allowing the calculation of durations.)

```{r}
library(tidyverse)
gait_download = read_csv("https://jlucasmckay.bmi.emory.edu/global/bmi510/gait.csv")
```
Note that there are usually two rows for each participant - one for the participant-level
information and one for the date of the recording. We can fill the dataset so that it is easier
to manipulate.

```{r}
gait_filled = gait_download |> fill(sex,dob,dw_icd)
view(gait_filled)
```
Note that this method just fills down! If you want to stop at the boundaries of a given participant,
you have to group.

```{r}
gait_filled = gait_download |>
  group_by(record_id) |>
  fill(sex,dob,dw_icd) |>
  ungroup()
```

Now that we have filled down, we can discard the extraneous patient-level rows.
```{r}
gait = gait_filled |>  filter(!is.na(redcap_repeat_instrument))
gait
```


Note that it is also nice to present some of the data as a well-formatted table.
We can do that with `slice_sample` and the `knitr` package.

```{r}
set.seed(510)
gait_sample = gait |>
  select(record_id,sex,speed,cadence) |> 
  na.exclude() |> 
  group_by(sex) |>
  slice_sample(n = 6)
gait_sample |> knitr::kable() |> kableExtra::kable_styling(full_width=F,position = "left")
```



## Joining datasets

You will end up doing a lot of `join`s when you work with medical data.
Here's a nice summary of how the different methods of joining work in the `dplyr` package: https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf

The ICD codes are not that easy to use, so let's join a dataset with a lookup table.
Most of the joins I do are `left_join`, which returns all rows from the first input and 
all columns from both inputs.

```{r}
icd = read_csv("https://jlucasmckay.bmi.emory.edu/global/bmi510/icd.csv")
```
There are too many classes here, so let's simplify. We can use `case_when` to recode this,
although it has a very curious syntax.

```{r}
icd$group = case_when(
grepl("agitans|arkinson",icd$diagnosis) ~ "PD",
grepl("sential",icd$diagnosis) ~ "ET",
T ~ "Other"
)
```

Now we can join this diagnostic information to the data.

```{r}
# gait = gait |> left_join(icd) # doesnt' work!
gait = gait |> rename(icd = dw_icd) |> left_join(icd)

gait2 = gait
gait2$icd = gait2$dw_icd
joined_data = left_join(gait2,icd)

```

Finally, we can reorder the columns for ease of use:

```{r}
gait = gait |> select(-starts_with("redcap")) |> select(-study_id) |> select(record_id, group, diagnosis, icd:sex, everything())

gait = gait |> arrange(group,dob)
gait
```

And make a plot.
```{r}
gait |>
  filter(group %in% c("PD","ET")) |> 
  ggplot(aes(x = speed, color = group)) + geom_density() +
  theme_minimal() +
  labs(color = "Study Group")
```


# Pivoting

The other hugely important topic for data wrangling is pivoting (mostly longer).
Let's look at this with some typical biomarker data.

```{r}
csf = read_csv("https://jlucasmckay.bmi.emory.edu/global/bmi510/csf.csv")
```

How are we going to plot this? Well, the first step is usually to totally melt the data:

```{r}
# how does column selection work?
csf_long = csf |> pivot_longer(-Group, values_to = "Expression", names_to = "Biomarker")
```

Although this is not the most easy way to think about the data, it does accommodate missing
data, different numbers of observations, etc. Once it is in this format, plotting is a snap.
We map Group to x, Expression to y, and Biomarker to the plot number.

```{r}
csf_long |> ggplot(aes(Group,Expression)) + geom_boxplot() + facet_wrap(~Biomarker, scales = "free_y") + theme_minimal()

csf_long |> ggplot(aes(Biomarker,Expression,fill=Biomarker)) + geom_boxplot() + facet_wrap(~Group, scales = "free_y") + theme_minimal()

```

# Making simple summary statistics

You can also group the melted data to create summary statistics pretty easily.
```{r}
sum_stats = csf_long |> group_by(Biomarker,Group) |> summarize(N = sum(!is.na(Expression)), Mean = mean(Expression, na.rm = T), SD = sd(Expression, na.rm = T))
sum_stats |> knitr::kable() |> kableExtra::kable_styling(full_width=F,position = "left")

```

```{r}
sum_stats |> write_csv("summary.csv",na="")
system("open summary.csv")
```




